### 2022-06-27 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@7fbeb69803b427454cd7d1bde2843f3d17331183 🚀

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South
America has the most evolutionary distinct hosts \emph{and} viruses,
followed by secondary hotspots in southeast Asia and the Rift Valley
region have mostly distinct viruses. Some degree of sampling bias may
contribute to these patterns: for example, South-America is one of the
places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{2,13,15} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{16}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there are
obvious and distinctive regional regimes of bat-betacoronavirus
diversification.

These global patterns recapitulate the evolutionary history of both the
order Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe bats
(Rhinolophidae) are both the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}) and the presumed ancestral host of
\emph{Betacoronavirus}.\textsuperscript{17} The hotspot of host richness
and viral diversity in southeast Asia---both of which are
disproportionately high considering the global landscape of bat species
richness---is almost entirely driven by dynamics in this
clade.\textsuperscript{3,14} In contrast, the Neotropical hotspot is
driven by a combination of distinct bat and virus adaptive radiation.
The hotspot of bat diversity in the Americas is driven by the New World
leaf-nosed bats (Phyllostomidae),\textsuperscript{18} a group of bats
that have undergone multiple explosive diversification events--- but
that are more often hosts of alphacoronaviruses. Out of the four main
groups of betacoronaviruses, only the subgenus \emph{Merbecovirus}---the
outgroup of the betacoronavirus tree---has been found in New World bats;
while found worldwide, the merbecoviruses are almost entirely restricted
to vesper bats (Vespertilionidae) {[}{]}. In this light, the hotspot of
virus distinctiveness in the Americas is the result of isolation by
vicariance--- a unique branch of coronavirus-bat coevolution, separate
from the rest of the globe.

@@

Despite differences in hotspots of evolutionary distinctiveness, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{19,20} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{21} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{22,23} in addition to being hardcoded in their
genome,\textsuperscript{24} may be related to a wide variety of
diets,\textsuperscript{21,25} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{26}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{27} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{28} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{29} and Europe\textsuperscript{30} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{31} Recent surveillance\textsuperscript{32}
has identified a novel \emph{Betacoronavirus} (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{33} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{34} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{16} stressing
the need for spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{19,20} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{35,36}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{37} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{38} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{39,40} and
vectors of pathogens that put a risk on human
health.\textsuperscript{41,42} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{13} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{14} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{17} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{43} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{14,16} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{44}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{45}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{46} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{47} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{48} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{49} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{50} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{51} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{52} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{53} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{54} LCBD works from a species-data
matrix (traditionally noted as \(\mathbf{Y}\)), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices\textsuperscript{55} transformed
it using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results by\textsuperscript{56} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{57} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{58} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{59} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{60} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{61} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{62} v1.6.12
with ModelFinder\textsuperscript{63} ultrafast bootstrap
approximation\textsuperscript{64} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{65} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

\leavevmode\hypertarget{ref-Villalobos2010DivFie}{}%
\CSLLeftMargin{18. }
\CSLRightInline{Villalobos, F. \& Arita, H. T. The diversity field of
New World leaf-nosed bats (Phyllostomidae). \emph{Global Ecology and
Biogeography} \textbf{19}, 200--211 (2010).}

\leavevmode\hypertarget{ref-VanBrussel2022ZooDis}{}%
\CSLLeftMargin{19. }
\CSLRightInline{Van Brussel, K. \& Holmes, E. C. Zoonotic disease and
virome diversity in bats. \emph{Current Opinion in Virology}
\textbf{52}, 192--202 (2022).}

>>>>

However, we find that the global pattern of betacoronavirus phylogenetic
distinctiveness is quite distinct from both bat host richness and
phylogenetic distinctiveness (fig.~\ref{fig:richness}; bottom). In
contrast to the sparsity of Neotropical betacoronavirus hosts, South
America has the most evolutionary distinct hosts \emph{and} viruses,
followed by secondary hotspots in southeast Asia and the Rift Valley
region have mostly distinct viruses. Some degree of sampling bias may
contribute to these patterns: for example, South-America is one of the
places where the fewest bat betacoronavirus sequences have been
generated,\textsuperscript{2,13,15} resulting in a sparser phylogenetic
tree, and artificially inflating distinctiveness; conversely,
disproportionate research effort in eastern China\textsuperscript{16}
may have led to a more complete inventory of the local diversity of
coronaviruses, again inflating these metrics relative to underlying
patterns. Even accounting for these potential biases, though, there is
obvious heterogeneity in betacoronavirus evolutionary distinctiveness.

These patterns recapitulate the evolutionary history of both the order
Chiroptera and the genus \emph{Betacoronavirus}. Horseshoe bats
(Rhinolophidae) are both the reservoirs of the SARS-like viruses
(subgenus \emph{Sarbecovirus}) and the presumed ancestral host of
\emph{Betacoronavirus}.\textsuperscript{17} The hotspot of host richness
and viral diversity in southeast Asia---both of which are
disproportionately high, considering the global landscape of bat species
richness---is almost entirely driven by adaptive radiation in this
clade\textsuperscript{3,14}. In contrast, the Neotropical hotspot of
viral distinctiveness is the result of isolation by vicariance for both
bats and their viruses. Out of the four main groups of
betacoronaviruses, only the subgenus \emph{Merbecovirus} has been found
in animals in the Americas---an introduction that is generally presumed
to be ancient. While comparatively understudied, New World
merbecoviruses have been found in the ghost-faced bats (Mormoopidae),
New World leaf-nosed bats (Phyllostomidae), and free-tailed bats
(Molossidae) (add cite: Olival 2020 PLoS Pathogens). The former two
groups are endemic to the Neotropics, while the latter two and
particularly the phyllostomids are responsible for the hotspot of bat
diversity in the Amazon. Together, these groups play host to a distinct
regime of bat-betacoronavirus coevolution in the Americas.

@@

Despite differences in hotspots of evolutionary distinctiveness, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of
congruence.\textsuperscript{18,19} In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat
species.\textsuperscript{20} Immune characteristics that allow bats to
be better adapted to infection by emerging
viruses,\textsuperscript{21,22} in addition to being hardcoded in their
genome,\textsuperscript{23} may be related to a wide variety of
diets,\textsuperscript{20,24} themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.\textsuperscript{25}

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses;\textsuperscript{26} MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases.\textsuperscript{27} For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti\textsuperscript{28} and Europe\textsuperscript{29} are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian
continents.\textsuperscript{30} Recent surveillance\textsuperscript{31}
has identified a novel \emph{Betacoronavirus} (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

Based on the previous result, we extracted the risk component from the
composite map (see Methods), to provide a single measure of risk varying
between 0 and 1. This measure is presented in fig.~\ref{fig:risk}. As
this map represents the potential risk, it must be weighed by the
potential for contacts with humans. As a proxy for this measure, we used
the proportion of build/urban land from the EarthEnv dataset: this is a
reasonable proxy for the density of humans per unit area, which
increases the probability of pathogen spread more
widely.\textsuperscript{32} Since human activity is required to amplify
the frequency of virus encounters and thus create areas of viral
amplification, mapping the potential risk against measures of land use
is required to generate a more actionable assessment of risk. This map
is presented in fig.~\ref{fig:compound}. Most of South America and
Europe are at comparatively lower risk, as although densely populated,
settlements tend to be in areas with lower potential risk. Regions like
Malaysia and the North coast of Australia have a high risk component,
but should represent a relatively lower effective risk due to low human
density. However, this mapping reveals that South-East Asia, the Indian
subcontinent, and parts of sub-Saharan Africa, are at high risk due to
the overlap between built areas and bat communities representing more
opportunities for cross-species transmission of betacoronaviruses. In
looking for the origins of SARS in China,\textsuperscript{33} present
serological evidence that strongest human-animal contact results in
higher risk of virus exposure, regardless of the animal species, but
that different types of contact had different impacts. Ideally,
finer-grained information about human activity (rather than human
presence through anthropisation) could allow to partition this risk
further, alebit at the cost of more hypotheses required to estimate the
amount of risk represented by each activity. Our map of purported high
risk/diversitifcation potential (Madagascar, South-America) overlay with
sampling gaps for \emph{Betacoronavirus},\textsuperscript{16} stressing
the need for spatially targeted monitoring and discovery.

@@

Bats are important reservoir hosts for different classes of
microorganisms, many of which a threat to human
health.\textsuperscript{18,19} Chiropterans emerged around 64 million
years ago and are one of the most diverse mammalian orders, with an
estimated richness of more than 1400 species.\textsuperscript{34,35}
They exhibit a broad variety of habitat use, behaviour, and feeding
strategies, putting them at key positions in the delivery and
provisioning of several ecosystem services, tied to important
ecosystem-derived benefits to human.\textsuperscript{36} For example,
bats are an essential component of many seed-dispersal
networks.\textsuperscript{37} Over two-thirds of bats are know to be
either obligate or facultative insectivores, therefore actively
contributing for agricultural pest control,\textsuperscript{38,39} and
vectors of pathogens that put a risk on human
health.\textsuperscript{40,41} Because bats are globally distributed and
have a long evolutionary history, phylogeographic and biogeographic
approaches are required to shed light on the contemporary distribution
of coevolutionary processes between bats and the pathogens they host.
Not all areas in which bats, viruses, and human are co-occuring are
facing a risk of spillover towards human populations, and the areas in
which this risk exist may not be facing risks of the same nature and
magnitude.

Here, we propose a simple freamework with broad explanatory power that
helps contextualize discoveries like highly divergent nobecoviruses in
Madagascar and the previously-neglected adaptive radiation of
sarbecoviruses outside of southern China and throughout southeast Asia.
In doing so, it advances ecological theory beyond the current state of
the art for global maps of emergence risk. For example, previous studies
that have used host richness as proxy have predicted a high diversity of
unsampled bat viruses,\textsuperscript{13} bat
coronaviruses,\textsuperscript{2} and even specifically
betacoronaviruses\textsuperscript{14} in both the Amazon and southeast
Asia. While we find that both regions are characterized by highly
divergent host and viral communities, our framework identifies key
differences between the regions. We find that Latin America is a hotspot
of both host and viral distinctiveness, suggesting that this branch of
the bat-betacoronavirus complex may be undergoing independent
evolutionary dynamics from the rest of the global pool, but with limited
potential for viral diversification--- a finding that is supported by
previous work indicating a higher rate of codivergence in Latin
America.\textsuperscript{2} In contrast, in southeast Asia, host
richness and viral distinctiveness are high but sharing is low; this
suggests a different type of evolutionary dynamics that could generate
high local diversity of viruses through host switching and viral
recombination (see \emph{e.g.},\textsuperscript{17} as well as the
discovery of recombinant viruses that share genetic material from both
the SARS-CoV and SARS-CoV-2 branches of the Sarbecovirus
lineage).\textsuperscript{42} Both of these regions are priority areas
for sampling, especially given predictions that they contain many bat
hosts of undiscovered betacoronaviruses.\textsuperscript{14,16} However,
both the evolutionary and ecological aspects of emergence risk are
likely higher in southeast Asia---a fact that will only become more
relevant, as bats track shifting climates and exchange viruses with
other species, creating a hotspot of cross-species transmission unique
to the region.\textsuperscript{43}

The diversity and diversification potential of bats responds to
anthropogenic factors others than shifting climates.\textsuperscript{44}
Land use changes could significantly decrease bat suitability, notably
through effects on diet and availability of
habitats.\textsuperscript{45} As our results establish that the
diversification of bats betacoronaviruses happens on top of processes
affecting hosts, biogeographic variation in human population density and
anthropogenic disturbances may feed into co-evolutionary dynamics.
Increase in humans-hosts contacts also increase the risk of emergence of
novel diseases,\textsuperscript{46} so does the changes in landscape
connectivity at local/regional scales.\textsuperscript{47} This
represents a challenge for both conservation strategies and disease
ecology: some areas can a high emergence risk and more potential for the
acquisition of zoonotic viruses through bat-human
encounters.\textsuperscript{48} In particular, the challenge ahead lies
in the need to quantify actual exposure (and risk) accounting for
several transmission scenarios, including both direct and indirect bat -
human interactions, and feeding back into the provision of ecosystem
services by bats.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN
data.\textsuperscript{49} The range maps were subsequently rasterized
using the \texttt{rasterize} function from
\texttt{GDAL}\textsuperscript{50} at a resolution of approximately
100kmx100km. For every pixel in the resulting raster where at least one
bat host of \emph{Betacoronavirus} was present, we extract the species
pool (list of all known bat hosts), which was used to calculate the
following risk assessment components: bat phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

@@

For every pixel, we measured Faith's Phylogenetic
Diversity\textsuperscript{51} based on a recent synthetic tree with
robust time calibration, covering about 6000 mammalian
species.\textsuperscript{52} Faith's PD measures the sum of unique
branches from an arbitrary root to a set of tips, and comparatively
larger values indicate a more phylogenetic diverse species pool. We
measured phylogenetic diversity starting from the root of the entire
tree (and not from Chiroptera); this bears no consequences on the
resulting values, since all branches leading up to Chiroptera are only
counted one per species pool, and (as we explain when describing the
assembly of the composite risk map), all individual risk components are
ranged in {[}0,1{]}. This measure incorporates a richness component,
which we chose not to correct for; the interpretation of the
phylogenetic diversity is therefore a weighted species richness, that
accounts for phylogenetic over/under-dispersal in some places.

@@

For every species pool, we measured its Local Contribution to
Beta-Diversity;\textsuperscript{53} LCBD works from a species-data
matrix (traditionally noted as \(\mathbf{Y}\)), where species are rows
and sites are columns, and a value of 1 indicates occurrence. We
extracted the Y matrix assuming that every pixel represents a unique
location, and following best practices\textsuperscript{54} transformed
it using Hellinger's distance to account for unequal bat richness at
different pixels. The correction of raw community data is particularly
important for two reasons: first, it prevents the artifact of richer
sites having higher importance; second, it removes the effect of overall
species richness, which is already incorporated in the phylogenetic
diversity component. High values of LCBD indicate that the pixel has a
community that is on average more dissimilar in species composition than
what is expected knowing the entire matrix, i.e.~a more unique
community. Recent results by\textsuperscript{55} shows that LCBD
measures are robust with regards to spatial scale, and are therefore
applicable at the global scale.

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap across
mammals.\textsuperscript{56} This network stores pairwise values of
viral community similarity. To project viral sharing values into a
single value for every pixel, we averaged the pairwise scores. High
values of the average sharing propensity means that this specific extant
bat assemblage is likely to be proficient at exchanging viruses.

@@

To visualize the aggregated risk at the global scale, we combine the
three individual risk components (phylogenetic diversity, compositional
uniqueness, and viral sharing) using an additive color
model.\textsuperscript{57} In this approach, every risk component gets
assigned a component in the RGB color model (phylogenetic diversity is
green, compositional uniqueness is red, and viral sharing is blue). In
order to achieve a valid RGB measure, all components are re-scaled to
the {[}0,1{]} interval, so that a pixel with no sharing, no phylogenetic
diversity, and no compositional uniqueness is black, and a pixel with
maximal values for each is white. This additive model conveys both the
intensity of the overall risk, but also the nature of the risk as colors
diverge towards combinations of values for three risk components. Out of
the possible combinations, the most risky in terms or rapid
diversification and spillover potential is high phylogenetic diversity
and low viral sharing,\textsuperscript{58} in that this allows multiple
independent host-virus coevolutionary dynamics to take place in the same
location. In the colorimetric space, this correspond to yellow --
because the HSV space is more amenable to calculations for feature
extraction,\textsuperscript{59} we measured the risk level by
calculating the angular distance of the hue of each pixel to a reference
value of 60 (yellow), and weighted this risk level by the value
component. Specifically, given a pixel with colorimetric coordinates
\((h,s,v)\), its ranged weighted risk value is

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent'',
``mutant'', ``GFP'', and ``recombinant''. We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using
MAFFT\textsuperscript{60} v1.4.0 (Algorithm FFT-NS-2, Scoring matrix
200PAM / k=2, gap open penalty 1.53m offset value 0.123) and a maximum
likelihood tree reconstructed in IQ-TREE\textsuperscript{61} v1.6.12
with ModelFinder\textsuperscript{62} ultrafast bootstrap
approximation\textsuperscript{63} with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

We first tested the hypothesis that hotspots of viral diversification
would track hotspots of bat diversification. To do so, we plotted the
number of known bat hosts (specifically only those included in the
phylogeny, so there was a 1:1 correspondence between data sources)
against the ``mean evolutionary distinctiveness'' of the associated
viruses. To calculate this, we derived the fair proportions evolutionary
distinctiveness\textsuperscript{64} for each of the viruses in the tree,
then averaged these at the bat species level, projected these values
onto their geographic distributions, and averaged across every bat found
in a given pixel. As such, this can be thought of as a map of the mean
evolutionary distinctiveness of the known viral community believed to be
associated with a particular subset of bats present.

@@

\leavevmode\hypertarget{ref-VanBrussel2022ZooDis}{}%
\CSLLeftMargin{18. }
\CSLRightInline{Van Brussel, K. \& Holmes, E. C. Zoonotic disease and
virome diversity in bats. \emph{Current Opinion in Virology}
\textbf{52}, 192--202 (2022).}

\leavevmode\hypertarget{ref-Letko2020BatVir}{}%
\CSLLeftMargin{19. }
\CSLRightInline{Letko, M., Seifert, S. N., Olival, K. J., Plowright, R.
K. \& Munster, V. J. Bat-borne virus diversity, spillover and emergence.
\emph{Nature Reviews Microbiology} \textbf{18}, 461--471 (2020).}

\leavevmode\hypertarget{ref-Banerjee2020NovIns}{}%
\CSLLeftMargin{20. }
\CSLRightInline{Banerjee, A. \emph{et al.} Novel Insights Into Immune
Systems of Bats. \emph{Frontiers in Immunology} \textbf{11}, 26 (2020).}
